---
title: "R for Non-R Users"
author: "Eugenio Paglino"
output: 
  revealjs::revealjs_presentation:
    transition: slide
    theme: simple
    center: false
---

<style type="text/css">
  .reveal p {
    text-align: left;
    font-size: 0.7em;
  }
  .reveal ul {
    display: block;
    font-size: 0.7em;
  }
  .reveal ol {
    display: block;
    font-size: 0.7em;
  }  
</style>

```{r setup, echo=F}

options(scipen = 999)

knitr::opts_chunk$set(echo = T,
                      message = F,
                      warning = F,
                      fig.align = 'center',
                      fig.height = 3,
                      fig.width = 5)

```

```{r, echo=F}

library(kableExtra)
library(bench)
library(tidyverse)

```

### What Is a Program?

In its essence, a program is a sequence of instructions that specifies how to perform a computation. 

Computations can be of different types:

- computing the mean of a set of numbers
- replacing the substring 'pre' with 'post' in a list of strings
- producing a graphic representation of some data

---

Every programming language has its own quirks but in general, every program (even the more complex ones) are a combination of a few basic instructions:

- **input**: Get data from the keyboard, a file, the network, or some other device.
- **output**: Display data on the screen, save it in a file, send it over the network, etc.
- **math**: Perform basic mathematical operations like addition and multiplication.
- **conditional execution**: Check for certain conditions and run the appropriate code.
- **repetition**: Perform some action repeatedly, usually with some variation.

## Coding Basics

The key building blocks of most languages are objects. In R, objects may be variables, arrays of numbers, character strings, functions, or more general structures built from such components. All R statements where you create objects, assignment statements, have the same form:

```{r, eval=F}
object_name <- value
```

Where `<-` is the assignment operator. This statement tells R that an object called `object_name` should be created and assigned the value `value`. A specific example could be:

```{r}
ourFirstString <- 'Hello World'
```

---

### Naming Conventions

It is a good practice to use descriptive names and a consistent style. People have their own taste. Some commonly used alternatives are:

```{r,eval=F}
iMostlyUseCamelCase
other_people_use_snake_case
some.people.use.periods
And_aFew.People_RENOUNCEconvention # Not recommended
```

You can inspect an object by typing its name:

```{r}
ourFirstString
```

---

### Types

Objects in R can be of different types (modes). The basic types of objects are:

- `numeric` (`integer` or `double`)
- `logical`
- `character`

A series of objects of the same type can be grouped into a `vector`. If you instead need to group objects of different types, you'll need to use `lists`. Dataframes, if you have head of them, are essentially a list of lists.

---

Objects can be easily converted from one type to another (when it makes sense to do so). Here is an example:

```{r}
(z <- 0:9) # Parentheses will make R print the object
```

```{r}
(digits <- as.character(z))
```

A further coercion reconstructs the numerical vector:

```{r}
(d <- as.integer(digits))
```

---

### Functions

A special and very useful type of object is a *function*. As a general concept, a function takes some objects as inputs or parameters and uses them to complete an action or produce another object. For example, the `seq()` function takes three main arguments and returns the sequence of values from `from` to `to` in steps of length `by`.

```{r}
seq(from=1, to=5, by=1)
```

---

You can define your own functions in the same way you create an object:

```{r,eval=F}
myFunction <- function(arg1 = val1, arg2 = val2, ...) {
  do_something(arg1,arg2)
}
```

For example, we can create a function that takes a vector of numbers and adds 2 to each of its elements:

```{r}
add2 <- function(numbers) {
  return(numbers + 2)
}

add2(1:4)
```

---

### Mathematical Operations

In our last example, you've seen how R can perform basic mathematical operations. Mathematical operators mostly work as expected:

```{r,eval=F}
 + # Addition 
 - # Subtraction 
 * # Multiplication 
 / # Division 
 ^ # Exponentiation 
 %% # Modulus
```
In addition to mathematical operators, R also has logical operators:

```{r,eval=F}
 & or && # AND 
 | or ||# OR 
 ! # NOT
```

---

### Conditional Execution

Often you need to perform one operation or another based on a given condition. This is what it is known as conditional execution. In R, you can implement it as follows:

```{r,eval=F}
if (someCondition) {
  
  executeThis()
  
} else {
  
  executeThat()
}
```

The expression or object that replaces `someCondition` must evaluate to either `TRUE` or `FALSE`.

---

Conditional expressions can also be chained:

```{r,eval=F}
if (someCondition) {
  executeThis()
} else if (someOtherCondition) {
  executeThat()
} else {
  executeSomethingElse()
}
```

or nested:

```{r,eval=F}
if (someCondition) {
  executeThis()
} else {
  if (someOtherCondition) {
    executeThat()
  }
}
```

---

### Loops and Repetition

Sometimes you need to repeat the same operation more than once with some variation. Here is an example:

```{r}
countdown <- function(n){
  while (n > 0) {
    print(n)
    n <- n - 1
  }
  print('Blastoff!')
}
```

The `while` statement check if a condition is `TRUE` (here `n>0`) and keeps executing the code in curly brackets when (if ever) it turns `FALSE`. Make sure that you are not setting up an infinite loop!

---

A second alternative when you know exactly how many iterations you need to run is the `for` statement. Here is how we wuld rewrite the `countdown` function with a `for` loop.

```{r}
countdown <- function(n){
  for (i in n:1) {
    print(i)
  }
  print('Blastoff!')
}
```

---

### A Primer on Code Performance

You might be tempted to use `for` loops to repeat an operation on all elements of a set of objects. That's because in R (and many other languages) iterations in a `for` loops are executed sequentially even when the could be executed in parallel.

R makes is relatively easy to transform certain `for` loops into their *vectorized* equivalent (i.e. implemented in such a way that multiple iterations are run simultaneously). Suppose we want to compute the exponential of log of `x`*0.3 + 1 for all the elements of a vector. We could define a function to perform such operation on a single number:

```{r}
arbitraryOperation <- function(number) {
  trasformedNumber <- log(number*0.3 + 1)
  trasformedNumber <- exp(trasformedNumber)
  return(trasformedNumber)
}
```

---

We could then apply this function to the elements of our vector with a `for` loop as follows:

```{r}
n <- 100
numbers <- seq(0,2,length.out=n)

result <- vector(mode = "list", length = n)
for(i in 1:n){
    result[[i]] = arbitraryOperation(numbers[i])
}
result <- as.numeric(result)

```

Alternatively, we could write a more compact expression using the `sapply` functional. This function applies a given function to all the elements of a vector and returns another vector. 

```{r, eval=F}
sapply(numbers,arbitraryOperation)
```

---

Because R is a smart language, many basic functions are automatically vectorized when applied over vectors. For example, the `exp`, the `+`, the `*`, and the `log` operations can all be applied directly to an entire vector (and our `arbitraryOperation` function inherits this property). Let us now compare the performance of these three alternatives.

```{r, echo=F}
mark(
  {
    result <- vector(mode = "list", length = n)
    for(i in 1:n){
      result[[i]] = arbitraryOperation(numbers[i])
    }
    result <- as.numeric(result)
  },
  sapply(numbers,arbitraryOperation),
  arbitraryOperation(numbers),
  iterations=100) %>%
  select(expression,median,mem_alloc) %>%
  mutate(expression=c('For Loop','Sapply','Vectorized')) %>%
  kable(col.names = c('Implementation','Median Execution Time','Memory Allocation')) %>%
  kable_styling()
  
```

With a vector of size 100, the `sapply` solution is about 30 times faster than the `for` loop and the vectorized version is about 29 times faster than the `sapply` solution. In general, if a vectorized version is available you should use it.

---

As a general note, you might not always care about performance. If you are not writing computationally intensive code, readability should be given a higher priority.

At the same time, R is not a fast language so if performance is key to your project, you might have to look [elsewhere](https://adv-r.hadley.nz/perf-improve.html#vectorise).

## A Brief Journey into the `tidyverse`

What we explored so far is part of the core R language. Indeed, except for the functionalities needed to time the code, we did not import any package. 

A package is a collection of functions pulled together by someone in order to perform a given set of operations. in R, packages are installed and loaded with the commands:

```{r,eval=F}
install.packages('packageName')
library(packageName)
```

Once a package is loaded, you can use its functions as if they were R core functions. If you want help wiht a function, you can type `?functionName` in our console to access its documentation.

---

There are many specialized packages in R, to build visualizations, to work with geographical data, to perform text analysis, etc.. Among R packages, the [`tidyverse`](https://www.tidyverse.org) collection stands out as coherent framework to perform all the most common data science operations, from reading the data, to cleaning it, to visualizing it.

Today, I'll be illustrating the following packages:

- `readr`: read and write data
- `dplyr` and `tidyr`: data wrangling
- `ggplot`: visualization
- `rmarkdown` and `knitr`: communicate

---

### `readr` Reading and Writing Data

`.csv` is by far the most common format for rectangular data. You can read `.csv` files into R as follows:

```{r,eval=F}
read_csv(pathToFile)
```

You can select which columns to read with the `col_select` option, specify column types with the `col_types` options, and specify a non standard delimiter with the `delim` option. 

`read_csv` is highly customizable and offers many useful options to take care of parsing problems. `readr` also offers the `read_fwf` function to read fixed width files (relatively common).

---

### `tidyr` and `dplyr` Transforming Data

Often, the data available to you is in a different format from the one you need. The `tidyverse` generally expects a dataset to follow these three principles:

There are three interrelated rules which make a dataset tidy:

1. Each variable must have its own column.
2. Each observation must have its own row.
3. Each value must have its own cell.

---

Consider the following table:

```{r, echo=F}
table2
```
This table violates rule 1 because two variables, `cases` and `population` share the same column.

---

We can easily convert the table to a tidy format with the following code:

```{r}
table2 %>%
  pivot_wider(names_from = 'type', values_from = 'count')
```

---

Imagine now that the data came as two separate tables:

```{r}
table4a # cases
table4b # population
```

---

Even in this case, we can easily create a tidy table. First, we need to collapse the different years into a unique variable:

```{r,results='hide'}
table4a %>%
  pivot_longer(`1999`:`2000`,
               names_to = 'year',values_to = 'cases')
table4b %>%
  pivot_longer(`1999`:`2000`,
               names_to = 'year',values_to = 'population')
```

---

and then we join them

```{r}
table4 <- table4a %>%
  pivot_longer(`1999`:`2000`,
               names_to = 'year',values_to = 'cases')
table4b %>% 
  pivot_longer(`1999`:`2000`,
               names_to = 'year',values_to = 'population') %>%
  left_join(table4,by=c('country','year'))
```

---

Other useful functions to transform datasets are:

- `separate`: to pull apart one column into multiple columns, by splitting wherever a separator character appears.
- `unite`: to combine multiple columns into a single column.
- `complete`: takes a set of columns, and finds all unique combinations. It then ensures the original dataset contains all those values, filling in where necessary.
- `fill`: takes a set of columns where you want missing values to be replaced by the most recent non-missing value.

---

Let us pause for a moment. In the code on the previous slides I've been frequently using the `%>%` operator. How does this work?

This pipe operator is implemented by the `magrittr` package and imported with the `tidyverse`. It passes the output of one operator or function to the next one as its first argument. Most `tidyverse` functions work very natural with the pipe but other functions work as well with the following syntax:

```{r,eval=F}
firstObject %>%
  someFunction(someArg=.,otherArg1=...,otherArg2=...)
```

---

In one of the previous examples, I used the `left_join` function to combine two datasets through a common *key*. This is a very common operation and is implemented by the `*_join` family of functions.

- `inner_join`: keeps only entries found in both databases
- `left_join`: keeps only entries found in left database
- `right_join`: keeps only entries found in the right database
- `outer_join`: keeps any entry found in one of the two databases

You can specify which keys should be used for the join operation with the `by` parameter.

```{r,eval=F}
table1 %>%
  *_join(table2,by=c('var1a'='var2a','var1b'='var)) # This is the most general syntax
```

---

You'll often want to add, remove, reorder, or modify variables in a dataset. This is the job of `dplyr`.

- You can add new variables with the `mutate` function (`transmute` removes exisiting variables).
- You can select which variables you want to keep (or remove) wih the `select` function.
- You can reorder variables with the `relocate` function.
- You can sort a dataset with the `arrange` function.

---

Here is an example:

```{r}
table1 %>%
  mutate(rate = cases/population) %>% # we add a new variables
  relocate(year) %>% # we move year to the first position
  select(country,year,rate) %>% # we keep only year, country, and rate
  arrange(year,country) # we sort by year and country
```

---

The syntax of `select` is used in many other function throughout the `tidyverse`. It is thus worth it to explore it a little bit more:

- `var1:var2`: Selects all variables between var1 and var 2
- `var1,var3,...,var5`: Selects individual variables var1, var3, ..., var5
- `starts_with('someString')`: Selets all variables starting with someString
- `ends_with('someString')`: Selets all variables ending with someString

The `-` operator in front of any selection (i.e. `select(-selection)`) excludes the selected variables and keeps the remaining ones

---

### Building Visualization with `ggplot2`

`ggplot2` is a general purpose library for visualizing data. The key principle of `ggplot2` is the idea of building up a graphic from multiple layers of data. You can learn more [here](http://vita.had.co.nz/papers/layered-grammar.pdf) but it's probably best to start with an example. We will be using the following classic dataset:

```{r}
head(iris)
```

---

We wish to plot sepal length against sepal width.

```{r}
ggplot(data=iris)+ # Notice that we use + and not the pipe %>%
  geom_point(mapping=aes(x=Sepal.Length,y=Sepal.Width))
```
We pipe the data to `ggplot` to set up the plot. Then we add our first layer `geom_point` and tell `ggplot2` that `Sepal.Length` should be mapped to the x-axis and `Sepal.Width` should be mapped to the y-axis.

---

Congratulations! This is your first graph. Suppose we want to change point shape based on the species.

```{r}
ggplot(data=iris) + # Notice that we use + and not the pipe %>%
  geom_point(mapping=aes(x=Sepal.Length,y=Sepal.Width,
                         shape=Species))
```

We just need to map the `Species` variable to the shape aesthetic.

---

If we felt that the shape alone is not enought to help the viewer separate points by species, we could add color:

```{r}
plot <- ggplot(data=iris,
               mapping=aes(x=Sepal.Length,y=Sepal.Width) )+
          geom_point(mapping=aes(shape=Species,color=Species)) 
plot
```

---

Suppose we want to focus on a specific portion of the graph:

```{r}
plot +
  coord_cartesian(xlim=c(6,7),ylim=c(2.5,3.5))
```

---

We can change the axis labels and add a title:

```{r}
plot +
  labs(x='Sepal Length',
       y='Sepal Width',
       title='The Relationship of Sepal Width and Lenght')
```

---

We could also add a regression line:

```{r} 
plot +
  geom_smooth(method = 'lm', se = F,color='gray') 
```

---

Maybe it would be better to have each species in a separate graph:

```{r, fig.width=7}
plot +
  facet_wrap(~Species)
```

---

We are not confined to points. Here is a histogram of sepal length.

```{r}
ggplot(iris) +
  geom_histogram(mapping=aes(x=Sepal.Length),
                 bins = 10,color='black',fill='white')
```

---

There are many different geometries you can use:

- Lines: `geom_line`
- Boxplots: `geom_boxplot`
- Densities: `geom_density`
- Maps: `geom_map`
- Error Bars: `geom_errorbar`
- Uncertainty Intervals: `geom_ribbon`

and many more.

---

You can customize almost every aspect of a plot: axes, labels, grid lines, legend, orientation, size, palettes. However, `ggplot2` also offers a set of themes that modify several aspects of a figure at once:

```{r} 
plot +
  theme_minimal()
```

---

```{r} 
plot +
  theme_bw()
```

---

```{r} 
plot +
  theme_classic()
```

and more are available through the `ggthemes` package.

---

### Communicating Your Results with RMarkdown

You might have noticed how this presentation combined formatted text, R code and plots. This presentation uses RMarkdown, a system within RStudio that lets you write code through a notebook interface and create reproducible documents.

If you clone the repository where I uploaded this presentation's material you'll be able to recreate this document with a simple click of the "Knit" button in RStudio.

You have many different output types you can choose from: html, markdown, doc, pdf (through LaTex), and others.

You can learn more about RMarkdown functionalities [here](https://rmarkdown.rstudio.com/lesson-1.html).